---
title: "Tidy Data with tidyr"
author: "Emma Grossman"
date: "4/21/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Tidy Data

```{r}
table1
```

```{r}
table2
```


Three rules make a dataset tidy:

1. each variable is a column
2. each observation has a row
3. each value has a cell

table1 is tidy, table2 is not. We can work with tidy data easily:
```{r}
# rate per 10,000
table1 %>%
  mutate(rate = cases/population*10000)
```

```{r}
# Cases per year
table1 %>%
  count(year, wt = cases)
```

```{r}
# Visualize changes over time
ggplot(table1, aes(year, cases))+
  geom_line(aes(group = country), color = "gray50")+
  geom_point(aes(color = country))
```


## Spreading and Gathering

There are usually two main problems: (1) one variable might be spread across multiple columns, or (2) one observations in multiple rows and bonus (3) both. Luckily, we have two functions in the `tidyverse` package, `gather()` and `spread()`.

### Gathering

`gather()` is helpful when column names are not the names of variables, but the values of other variables. For example, in `table4a`, the column names are the values of the `year` variable.

```{r}
table4a
```

```{r}
table4a <- table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases")

table4b <- table4b %>%
  gather(`1999`, `2000`, key = "year", value = "population")
```

We can combine these two datasets into one to create our tidy data:
```{r}
left_join(table4a, table4b)
```

### Spreading

Spreading is the opposite of gathering. An observation is scattered across multiple rows, like `table2`.
```{r}
table2
```


```{r}
spread(table2, key = type, value = count)
```

## Separating and Pull

`table3` has yet another problem. Two variables are in one column of data.
```{r}
table3
```

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"))
```

By default, it separates where it doesn't see an alphanumeric character, like the forward slash above. We can also specify if need be:
```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/")
```

If we look at variables `cases` and `population`, though, they are character vectors instead of numeric vectors, so we should change that. 
```{r}
table3 %>%
  separate(rate, 
           into = c("cases", "population"), 
           convert = TRUE)
```

`convert` attempts to change the variable type into more fitting ones, if there are any. 

If we pass an integer into `sep`, it is interpreted as a position. For example,
```{r}
table3 %>%
  separate(year, into = c("century", "year"), sep = 2)
```

### Unite

Unite is the opposite of `separate`: it joins two columns.

```{r}
table5 %>%
  unite(new, century, year)
```

If we want to remove the underscore, we can.
```{r}
table5 %>%
  unite(new, century, year, sep = "")
```

## Missing Values

We have have explicitly missing data, flagged with `NA` or implicitly missing data, which just isn't present.

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```


We can make the implicit missing values explicit:
```{r}
stocks %>%
  spread(year, return)
```

Or, if our missing data isn't important, we can make it implicit:
```{r}
stocks %>%
  spread(year, return) %>%
  gather(year, return, `2015`:`2016`, na.rm = TRUE)
```

We can also use `complete()`, which finds all unique combinations of variables:
```{r}
stocks %>%
  complete(year, qtr)
```

Another tool for missing variables is utilized when missing values indicate that the previous value should be carried forward:
```{r}
treatment <- tribble(
  ~person,           ~treatment, ~response,
  "Derrick Whitmore", 1,         7,
  NA,                 2,         10,
  NA,                 3,         9,
  "Katherine Burke",  1,         4
)
```


This can be solved with `fill()`.
```{r}
treatment %>%
  fill(person)
```

# Case Study
```{r}
tidyr::who
```


Who is messy.

```{r}
who1 <- who %>%
  gather(
    new_sp_m014:newrel_f65, key = "key",
    value = cases,
    na.rm = TRUE
  )
who1
```

```{r}
who1 %>%
  count(key)
```


```{r}
who2 <- who1 %>%
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))
```

```{r}
who3 <- who2 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```

```{r}
who3 %>%
  count(new)

who4 <- who3 %>%
  select(-new, -iso2, -iso3)
```


```{r}
who5 <- who4 %>%
  separate(sexage, c("sex", "age"), sep = 1)
who5
```

















