---
title: "Strings with stringr"
author: "Emma Grossman"
date: "5/12/2021"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
```

The focus of this chapter is regex (I have experience with this from Computational Linguistics). 

## String Basics

```{r}
string1 <- "This is a string"
string2 <- 'To put a "quote" inside a string, use single quotes'
```

We can use a backslash to "escape" a string:
```{r}
double_quote <- "\""
single_quote <- '\''
```

So, to include a literal backslash, we need to double it.
```{r}
x <- c("\"", "\\")
x
writeLines(x)
```

Sometimes an odd string refers to a non-English character:
```{r}
(x <- "\u00b5")
plot(runif(10), runif(10), xlab ="\u00b5")
# a way to have Greek letters on our plots without the package
# I used for my MS project
```

## String Length

Functions in the `stringr` generally begin with the prefix `str_`.

```{r}
str_length(c("a", "R for data science", NA))
```


## Combining Strings

```{r}
str_c("x", "y")
str_c("x", "y", "z")
str_c("x", "y", sep = ", ")
```

Missing values are contagious.
```{r}
x <- c("abc", NA)
str_c("|-", x, "-|")
str_c("|-", str_replace_na(x), "-|")
```

As shown above, it is a vectorized function and recycles strings:
```{r}
str_c("prefix-", c("a", "b", "c"), "-suffix")
```

Objects of length 0 are dropped, which is useful with `if`.
```{r}
name <- "Emma"
time_of_day <- "morning"
birthday <- FALSE

str_c(
  "Good ", time_of_day, " ", name,
  if(birthday) "and HAPPY BIRTHDAY",
  "."
)
```

We can collapse a vector of strings into just one with an argument:
```{r}
str_c(c("x", "y", "z"), collapse = ", ")
str_c(c("x", "y", "z"), collapse = "")
```

## Subsetting strings

```{r}
# str_sub(vector, start, end)
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)

# negative numbers count backward
str_sub(x, -3, -1)
```

If we try to subset with more characters than exist, the funciton will return as many characters as it can:
```{r}
str_sub("a", 1, 5)
```

Freaking cool:
```{r}
str_sub(x, 1, 1) <- str_to_lower(str_sub(x, 1, 1))
x
```

## Locales

Changing from upper to lower case or vise versa can be complicated because different languages use different rules to do so. The default Locale used by R depends on the current locale provided by the operating system of the computer. 

Sometimes how letters are storted varies by region:
```{r}
x <- c("apple", "eggplant", "banana")
str_sort(x, locale = "en") # english
str_sort(x, locale = "haw") # hawiian
```


## Matching Patterns with Regular Expressions

### Basic Matches

Exact match:
```{r}
x <- c("apple", "banana", "pear")
str_view(x, "an")
```

`.` matches any character (except a new line)
```{r}
str_view(x, ".a.")
```

Use a double backslash to escape using `.`'s special function and ask for an exact match.
```{r}
dot <- "\\."
writeLines(dot)
str_view(c("abc", "a.c", "bef"), "a\\.c")
```

To identify a literal backslash, we need four of them.
```{r}
x <- "a\\b"
writeLines(x)
str_view(x, "\\\\")
```

### Anchors 
 `^` matches the start of a string 
 `$` matches the end of a string
 
```{r}
x <- c("apple", "banana", "pear")
str_view(x, "^a")
str_view(x, "a$")
```
 
We can force regex to only match a complete string by sandwiching it between the anchors:
```{r}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")
str_view(x, "^apple$")
```

### Character Classes and Alternatives

- \\d matches any digit
- \\s matches whitespace (space, tab, newline)
- [abc] matches a, b or c
- [^abc] maches anything except a, b or c

```{r}
str_view(c("gray", "grey"), "gr(e|a)y")
```

### Repetition

How many times a pattern is matched.

- ?: 0 or 1
- +: 1 or more
- *: 0 or more

```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"
str_view(x, "CC?")
str_view(x, "CC+")
str_view(x, "C[LX]+")
```

We can utilize numbers as well.
- \{n\}: exactly n times
- \{n,\}: n or more times
- \{,m\}: at most m times
- \{n,m\}: between n and m times

```{r}
str_view(x, "C{2}")
str_view(x, "C{2,}")
str_view(x, "C{2,3}")
```

By default, the longest match possible will always be returned, but this can be changed by adding a `?` after them.
```{r}
str_view(x, "C{2,3}?")
```

Regex is powerful but complex, and it generally should not be the first method to use when trying to solve a problem. It can be more trouble than it is worth.

### Detect Matches

`str_detect` determines if a character expression matches a pattern.
```{r}
x <- c("apple", "banana", "pear")
str_detect(x, "e")
```

```{r}
# How many common words start with t?
sum(str_detect(words, "^t"))

# What proportion of common words end with a vowel?
mean(str_detect(words, "[aeiou]+$"))
```

### Exact Matches

```{r}
length(sentences)
head(sentences)
```

```{r}
colors <- c(
  "red", "orange", "yellow", "greeen", "blue", "purple"
)
color_match <- str_c(colors, collapse = "|")
color_match
```

```{r}
has_color <- str_subset(sentences, color_match)
matches <- str_extract(has_color, color_match)
head(matches)
```


```{r}
more <- sentences[str_count(sentences, color_match) > 1]
str_view_all(more, color_match)
```


### Grouped Matches

```{r}
noun <- "(a|the) ([^ ]+)"
has_noun <- sentences %>%
  str_subset(noun) %>%
  head(10)

has_noun %>%
  str_extract(noun)
```

We could also do this with `tibble`:
```{r}
tibble(sentence = sentences) %>%
  tidyr::extract(
    sentence, c("article", "noun"), "(a|the) ([^ ]+)",
    remove = FALSE
  )
```


### Replacing Matches

```{r}
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")
str_replace_all(x, "[aeiou]", "-")
```

```{r}
x <- c("1 house", "2 cars", "3 people")
str_replace_all(x, c("1" = "one", "2" = "two", "3" = "three"))
```

```{r}
sentences %>%
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") %>%
  head(5)
```

### Splitting 

```{r}
sentences %>%
  head(5) %>%
  str_split(" ")
```

```{r}
sentences %>%
  head(5) %>%
  str_split(" ", simplify = TRUE)
```


```{r}
x <- "This is a sentence. This is another sentence."
str_view_all(x, boundary("word"))
```










