---
title: 'Advanced R: Vectors'
author: "Emma Grossman"
date: "6/1/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 3.2 Atomic vectors

There are four types of atomic vectors: logical, integer, double, and character. Two other very rare types exist: complex and raw. 

`c()` is short for combine.

We can use `typeof()` to determine the type of a vector.

```{r}
lgl_var <- c(TRUE, FALSE)
int_var <- c(1L, 6L, 10L)
dbl_var <- c(1, 2.5, 4.5)
chr_var <- c("these are", "some strings")

typeof(lgl_var)
```

`NA` is used for missing data. It is infections: most operations including a `NA` value returns another `NA` value, with a few exceptions.

```{r}
NA > 5
10 * NA
!NA
```

The exceptions:
```{r}
NA ^ 0
NA | TRUE
NA & FALSE
```

We use `is.na()` to test for missingness rather than `something == NA`.

If we attempt to combine vectors of different types, coercion occurs. 

```{r}
str(c("a", 1))
```

```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)

# Total number of TRUEs
sum(x)

# Proportion that are TRUE
mean(x)
```

```{r}
as.integer(c("1", "1.5", "a"))
```

## 3.3 Attributes

Matrices, arrays, factors, and date-times are data structures built on vectors using attributes.

Attributes attach metadata to data.

```{r}
a <- 1:3
attr(a, "x") <- "abcdef"
attr(a, "x")

attr(a, "y") <- 4:6
str(attributes(a))

# Or equivalently
a <- structure(
  1:3, 
  x = "abcdef",
  y = 4:6
)
str(attributes(a))
```

Only two attributes are maintained: **names** and **dim**

Three ways to create names are:
```{r}
# When creating it: 
x <- c(a = 1, b = 2, c = 3)

# By assigning a character vector to names()
x <- 1:3
names(x) <- c("a", "b", "c")

# Inline, with setNames():
x <- setNames(1:3, c("a", "b", "c"))
```


We can remove names with `x <- unname(x)` or with `names(x) <- NULL`.

```{r}
str(1:3)                   # 1d vector
str(matrix(1:3, ncol = 1)) # column vector
str(matrix(1:3, nrow = 1)) # row vector
str(array(1:3, 3))         # "array" vector
```

## 3.4 S3 atomic vectors

> One of the most important vector attributes is class, which underlies the S3 object system. Having a class attribute turns an object into an S3 object, which means it will behave differently from a regular vector when passed to a generic function. Every S3 object is built on top of a base type, and often stores additional information in other attributes.

There are four important S3 atomic vectors in R: factor vectors, date vectors, date-time vectors, and duration (difftime) vectors. 

Factor vectors must only contain a set of predefined values and they are built from integer vectors (with class and levels).

```{r}
x <- factor(c("a", "b", "b", "a"))
x

typeof(x)
attributes(x)
```

```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

There is a subset of factors called ordered factors, in which the levels of the factor have an inherent ordering. Low, medium, and high, for example.

```{r}
grade <- ordered(c("b", "b", "a", "c"), levels = c("c", "b", "a"))
grade
```

Date vectors derive from double vectors and they have an attribute class of date.

```{r}
today <- Sys.Date()

typeof(today)
attributes(today)
```

```{r}
date <- as.Date("1980-02-01")
unclass(date)
```

When we unclass a date, it provides the number of days since 1970-01-01.

A duration vector is built on a double, with a `unit` attribute.

```{r}
one_week_1 <- as.difftime(1, units = "weeks")
one_week_1

typeof(one_week_1)
attributes(one_week_1)

one_week_2 <- as.difftime(7, units = "days")
one_week_2

typeof(one_week_2)
attributes(one_week_2)
```

## 3.5 Lists

Lists are complex because elements can be of any type (*technically* elements of a list are the same type because they are references).

```{r}
l1 <- list(
  1:3, 
  "a", 
  c(TRUE, FALSE, TRUE), 
  c(2.3, 5.9)
)

typeof(l1)

str(l1)
```

Lists are often smaller than we think they are.
```{r}
lobstr::obj_size(mtcars)

l2 <- list(mtcars, mtcars, mtcars, mtcars)
lobstr::obj_size(l2)
```

Lists can contain other lists:
```{r}
l3 <- list(list(list(1)))
str(l3)
```

We can use dimension to create list-matrices and list-arrays

```{r}
l <- list(1:3, "a", TRUE, 1.0)
dim(l) <- c(2, 2)
l

l[[1, 1]]
```

## 3.6 Data frames and tibbles

Data frames are a named list vector with attributes `names`, `row.names` and a class of `data.frame`.

```{r}
df1 <- data.frame(x = 1:3, y = letters[1:3])
typeof(df1)

attributes(df1)
```

Each column of a data frame is restricted to be the same length.

A Tibble is a modern reimagination of the data frame.

> A concise, and fun, way to summarise the main differences is that tibbles are lazy and surly: they do less and complain more. 

```{r}
library(tibble)

df2 <- tibble(x = 1:3, y = letters[1:3])
typeof(df2)

attributes(df2)
```

```{r}
df <- data.frame(
  x = 1:3, 
  y = c("a", "b", "c")
)
str(df)
```

```{r}
df1 <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE
)
str(df1)
```

Tibbles will never coerce their input.
```{r}
df2 <- tibble(
  x = 1:3, 
  y = c("a", "b", "c")
)
str(df2)
```

Data frames will change the name of columns that are non-syntactic, tibbles do not.

```{r}
names(data.frame(`1` = 1))

names(tibble(`1` = 1))
```

If an element of a data frame or tibble has a shorter length, it will be recycled. Data frames recycle for any shorter length variable, but tibbles will only recyle if a the shorter variable is of length 1. 
```{r error = TRUE}
data.frame(x = 1:4, y = 1:2)
data.frame(x = 1:4, y = 1:3)

tibble(x = 1:4, y = 1)
tibble(x = 1:4, y = 1:2)
```

One final difference is that tibbles allow the user to create new variables using newly created variables, for example:

```{r}
tibble(
  x = 1:3,
  y = x * 2 # x was just created
)
```

We can have row names.
```{r}
df3 <- data.frame(
  age = c(35, 27, 18),
  hair = c("blond", "brown", "black"),
  row.names = c("Bob", "Susan", "Sam")
)
df3
```


```{r}
rownames(df3)

df3["Bob", ]
```

Matrices are transposeable, data frames are not. It is best to avoid row names, they are an added complication.

```{r}
df3[c(1, 1, 1), ]
```

So, tibbles do not support row names.

```{r}
as_tibble(df3, rownames = "name")
```

```{r}
dplyr::starwars
```

I've definitely run into this issue before:
> When you subset columns with df[, vars], you will get a vector if vars selects one variable, otherwise you’ll get a data frame. This is a frequent source of bugs when using [ in a function, unless you always remember to use df[, vars, drop = FALSE].

Another inconvenient subsetting behavior:
> When you attempt to extract a single column with df$x and there is no column x, a data frame will instead select any variable that starts with x. If no variable starts with x, df$x will return NULL. This makes it easy to select the wrong variable or to select a variable that doesn’t exist.

Tibbles modify this by only returning tibbles and they don't do partial matching.

```{r error = TRUE}
df1 <- data.frame(xyz = "a")
df2 <- tibble(xyz = "a")

str(df1$x)
str(df2$x)
```

Data frames can contain lists.

```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)

data.frame(
  x = 1:3, 
  y = I(list(1:2, 1:3, 1:4))
)
```

Tibbles make this a little easier:
```{r}
tibble(
  x = 1:3, 
  y = list(1:2, 1:3, 1:4)
)
```

It is possible to have a matrix or array as column of data.

```{r}
dfm <- data.frame(
  x = 1:3 * 10
)
dfm$y <- matrix(1:9, nrow = 3)
dfm$z <- data.frame(a = 3:1, b = letters[1:3], stringsAsFactors = FALSE)

str(dfm)
dfm[1, ]
```

## 3.7 `NULL`

> NULL is special because it has a unique type, is always length zero, and can’t have any attributes:

```{r error = TRUE}
typeof(NULL)

length(NULL)

x <- NULL
attr(x, "y") <- 1
```

```{r}
is.null(NULL)
```







