---
title: "Pipes with magrittr"
author: "Emma Grossman"
date: "5/26/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The pipe character `%>%` comes from Stefan Milton's package, `magrittr`. The package isn't loaded automatically, but the character is.

```{r}
library(tidyverse)
library(magrittr)
```

Big picture: the pipe character helps to make code easily writable and readable. 

```{r}
diamonds <- ggplot2::diamonds
diamonds2 <- diamonds %>%
  dplyr::mutate(price_per_carat = price / carat)
```

Using `magrittr`: 
```{r eval = FALSE}
foo_foo %>%
  hop(through = forest) %>%
  scoop(up = field_mice) %>%
  bop(on = head)
```


Behind the scenes, `magrittr` does something like this:
```{r eval = FALSE}
my_pipe <- function(.) {
  . <- hop(., through = forest)
  . <- scoop(., up = field_mice)
  bop(., on = head)
}
my_pipe(foo_foo)
```

The pipe doesn't work with functions that utilize the current environment, like `assign()`, `get()`, and `load()`, or functions that use lazy evaluation, like `tryCatch()`. 

There is also the "tee" pipe, `%T>%`, which returns the lefthand side rather than the righthand. 

```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()

rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```














