---
title: 'Week 3: Tibbles with tibble'
author: "Emma Grossman"
date: "4/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

tibbles are data frames that tweak outdated behaviors of data frames. 

## Creating Tibbles

Coerce a data frame into a tibble:
```{r}
as_tibble(iris)
```


Create a new tibble from individual vectors:
```{r}
tibble(
  x = 1:5,
  y = 1,
  z = x^2 + y
)
```

We can see that tibble recycles inputs of length one, like y in the example above. 

tibble never changes the type of input (converts strings to factors), changes the names of variables, or creates row names. tibble can have unconventional or *nonsyntactic* variable names, however:

```{r}
tibble(
  `:)` = "smile",
  ` ` = "space",
  `2000` = "number"
)
```

We can also create a tibble with `tribble()`, which stands for transposed tibble; it is customized for data entry in code.
```{r}
tribble(
  ~x, ~y, ~z, # column names have a tilde in front
  #--/--/----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```


## Tibbles Versus data.frame

The main differences are printing and subsetting.

### Printing

When printing a tibble we only see the first 10 rows and all columns that fit on the screen. It also reports column type.

```{r}
tibble(
  a = lubridate::now()   + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```


tibbles are designed so we don't overwhelm our console.

We can change the number of rows and `width` of the display:
```{r}
nycflights13::flights %>%
  print(n = 11, width = Inf)
```

We can use base R and the pipe to view a dataset in a new tab. This is helpful after a long string of manipulations.
```{r}
# nycflights13::flights %>%
#   View()
```



### Subsetting

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
df[["x"]]

# Extract by position
df[[1]]
```

If we want to use the pipe, there is a special placeholder.
```{r}
df %>% .$x
df %>% .[["x"]]
```

We can also convert back to a data.frame with `as.data.frame()`.
```{r}
class(as.data.frame(df))
```


```{r}
# ?enframe
enframe(1:3)
enframe(list(one = 1, two = 2:3, three = 4:6))
```











