---
title: "Factors with forcats"
author: "Emma Grossman"
date: "5/19/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forcats)
```

Factors are used for categorical variables and were historically must easier to work with than characters. For this reason, base R will sometimes automatically convert strings to factors: `stringsAsFactors` style.

# Creating Factors

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
x2 <- c("Dec", "Apr", "Jam", "Mar")
```

Using a string could result in: (1) typos and (2) it doesn't sort in a useful way.

```{r}
sort(x1)
```

These problems can be fixed with a factor.
```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```


```{r}
y2 <- factor(x2, levels = month_levels)
y2

y2 <- parse_factor(x2, levels = month_levels)
```


```{r}
# alphabetical
factor(x1)
```

```{r}
# order of appearance
f1 <- factor(x1, unique(x1))
f1

# order of appearance (after the fact)
f2 <- x1 %>% factor %>% fct_inorder()
f2
```


## General Social Survey

```{r}
gss_cat
```


```{r}
gss_cat %>%
  count(race)

ggplot(gss_cat, aes(race))+
  geom_bar()
```

```{r}
# ggplot drops levels w/o values but we can force them to display

ggplot(gss_cat, aes(race))+
  geom_bar()+
  scale_x_discrete(drop = FALSE)
```


## Modifying Factor Order

```{r}
relig <- gss_cat %>%
  group_by(relig) %>%
  summarize(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig, aes(tvhours, relig)) + geom_point()
```

```{r}
ggplot(relig, aes(tvhours, fct_reorder(relig, tvhours))) + geom_point()
```


```{r}
relig %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig))+
  geom_point()
```

```{r}
rincome <- gss_cat %>%
  group_by(rincome) %>%
  summarize(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(rincome, aes(age, fct_reorder(rincome, age)))+
  geom_point()
```

Not a good idea, because there is an inherent ordering to income. But, we may want to pull not applicable to the front.

```{r}
ggplot(rincome, aes(age, fct_relevel(rincome, "Not applicable")))+
  geom_point()
```

> `fct_reorder2()` reorders the factor by the y values associated with the largest x values. This makes the plot easier to read because the line colors line up with the legend:

```{r}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  count(age, marital) %>%
  group_by(age) %>%
  mutate(prop = n / sum(n))

ggplot(by_age, aes(age, prop, colour = marital)) +
  geom_line(na.rm = TRUE)

ggplot(by_age, aes(age, prop, colour = fct_reorder2(marital, age, prop))) +
  geom_line() +
  labs(colour = "marital")
```


```{r}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(marital))+
  geom_bar()
```


## Modifying Factor Levels

`fct_recode()` is super helpful.

```{r}
gss_cat %>%
  count(partyid)
```


```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
  )) %>%
  count(partyid)
```

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party"
  )) %>%
  count(partyid)
```


```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```


```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig)) %>%
  count(relig)
```


```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n = 10)) %>%
  count(relig, sort = TRUE) %>%
  print(n = Inf)
```

















