---
title: "Dates and Times with lubridate"
author: "Emma Grossman"
date: "5/24/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(nycflights13)
```

## Creating Date/Times

When looking at date/time data we have three options (1) a date, (2) a time, or (3) a date-time. R doesn't have a base class for storing times, so the `hms` packages functions well. 

Good practice is to always uses the simplest measure of time as possible. 

```{r}
today()
now()
```

### From Strings

A date/time generally is created via a string, from individual date-time components or from existing date/time object. 

```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
ymd(20170131)
```

The above functions create dates, but we can also create date/times. 

```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")
```

> You can also force the creation of a date-time from a date by supplying a time zone:

```{r}
ymd(20170131, tz = "UTC")
```

### From Individual Components

```{r}
flights %>%
  select(year, month, day, hour, minute)
```

We can combine all of these elements into a date/time.

```{r}
flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(
    departure = make_datetime(year, month, day, hour, minute)
  )
```



```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt
```


```{r}
flights_dt %>%
  ggplot(aes(dep_time))+
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day
```


```{r}
flights_dt %>%
  filter(dep_time < ymd(20130102)) %>%
  ggplot(aes(dep_time)) +
  geom_freqpoly(binwidth = 600) # 10 minutes
```

For date-times, 1 = 1 second. For dates, 1 = 1 day. 

### From Other Types

We can switch between dates and date/times.
```{r}
as_datetime(today())
as_date(now())
```

## Date-Time Components

### Getting Components

We can extract components from a date-time object.
```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
mday(datetime) # day of the month
yday(datetime) # day of the year
wday(datetime) # day of the week

month(datetime, label = TRUE)
wday(datetime, label = TRUE, abbr = FALSE)
```


```{r}
flights_dt %>%
  mutate(wday = wday(dep_time, label = TRUE)) %>%
  ggplot(aes(x = wday)) +
  geom_bar()
```

```{r}
flights_dt %>% 
  mutate(minute = minute(dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, avg_delay)) +
    geom_line()
```


```{r}
sched_dep <- flights_dt %>% 
  mutate(minute = minute(sched_dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n())

ggplot(sched_dep, aes(minute, avg_delay)) +
  geom_line()
```


```{r}
ggplot(sched_dep, aes(minute, n)) +
  geom_line()
```

### Rounding

`floor_date()`, `round_date()`, and `ceiling_date()` can be used to round a date to a nearby time.

```{r}
flights_dt %>% 
  count(week = floor_date(dep_time, "week")) %>% 
  ggplot(aes(week, n)) +
    geom_line()
```

### Setting Components

```{r}
(datetime <- ymd_hms("2016-07-08 12:34:56"))

year(datetime) <- 2020
datetime
month(datetime) <- 01
datetime
hour(datetime) <- hour(datetime) + 1
datetime
```

```{r}
update(datetime, year = 2020, month = 2, mday = 2, hour = 2)
```

```{r}
ymd("2015-02-01") %>% 
  update(mday = 30)
ymd("2015-02-01") %>% 
  update(hour = 400)
```

```{r}
flights_dt %>% 
  mutate(dep_hour = update(dep_time, yday = 1)) %>% 
  ggplot(aes(dep_hour)) +
    geom_freqpoly(binwidth = 300)
```

## Time Spans

- **durations** - exact number of seconds
- **periods** - human units like weeks and months
- **intervals** - starting and ending point

### Durations

When we subtract two dates, we get a difftime object.
```{r}
today() - ymd(19960905) # how old in days

as.duration(today() - ymd(19960905))
```

```{r}
dseconds(15)
dminutes(10)
dhours(c(12, 24))
ddays(0:5)
dweeks(3)
dyears(1)
```

```{r}
2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)
```

```{r}
tomorrow <- today() + ddays(1)
last_year <- today() - dyears(1)
```


```{r}
one_pm <- ymd_hms("2016-03-12 13:00:00", tz = "America/New_York")

one_pm
one_pm + ddays(1)
```

### Periods

```{r}
one_pm
one_pm + days(1)
```

```{r}
seconds(15)
minutes(10)
hours(c(12, 24))
days(7)
months(1:6)
weeks(3)
years(1)
```

```{r}
10 * (months(6) + days(1))
days(50) + hours(25) + minutes(2)
```

```{r}
# A leap year
ymd("2016-01-01") + dyears(1)
ymd("2016-01-01") + years(1)

# Daylight Savings Time
one_pm + ddays(1)
one_pm + days(1)
```

```{r}
flights_dt %>% 
  filter(arr_time < dep_time)
```


```{r}
flights_dt <- flights_dt %>% 
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight * 1),
    sched_arr_time = sched_arr_time + days(overnight * 1)
  )
```

```{r}
flights_dt %>% 
  filter(overnight, arr_time < dep_time) 
```

### Intervals

```{r}
years(1) / days(1)
```

```{r}
next_year <- today() + years(1)
(today() %--% next_year) / ddays(1)
```

```{r}
(today() %--% next_year) %/% days(1)
```

### Time Zones
```{r}
Sys.timezone()
```

```{r}
(x1 <- ymd_hms("2015-06-01 12:00:00", tz = "America/New_York"))
(x2 <- ymd_hms("2015-06-01 18:00:00", tz = "Europe/Copenhagen"))
(x3 <- ymd_hms("2015-06-02 04:00:00", tz = "Pacific/Auckland"))
```

```{r}
x1 - x2
x1 - x3
```

```{r}
x4 <- c(x1, x2, x3)
x4
```

