---
title: "Relational Data with dplyr"
author: "Emma Grossman"
date: "4/26/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In the real world, it is rare that data analysis only involves a single table of data. Multiple tables are called relational data, because the relationships between tables are as important as the individual datasets.

There are several ways to join data from multiple tables: 

- *mutating joins* - add new variables to a df from matching observations in another
- *filtering joins* - filter observations from one df based on whether they match an observation in the other table
- *set operations* - treat observations as if they were set elements

```{r}
library(tidyverse)
library(nycflights13)
```

## nycflights13

There are four tibbles related to flights.
```{r}
head(airlines)
head(airports)
head(planes)
head(weather)
```

`flights` and `planes` have the variables `tailnum` in common; `flights` and `airlines` have `carrier` in common; `flights` and `airports` have `origin` and `dest` in common; and `flights` and `weather` have `origin` in common.

### Keys

*Keys* are variables used to connect pairs of tables. A *primary key* uniquely identifies  an observation in a table, like an ID column. A *foreign key* uniquely identifies an observation in another table.

One way to ensure that a primary key indeed uniquely identifies each observation is by using `count()`.
```{r}
planes %>%
  count(tailnum) %>%
  filter(n > 1)
```

Since there are no `tailnum` assigned to multiple planes, it is indeed unique.
```{r}
weather %>%
  count(year, month, day, hour, origin) %>%
  filter(n > 1)
```

Sometimes there is no primary key. No combination of the variables uniquely identifies each observation. 
```{r eval = FALSE}
flights %>%
  count(year, month, day, flight) %>%
  filter(n > 1)
```

What if we include `tailnum`?
```{r}
flights %>%
  count(year, month, day, tailnum) %>%
  filter(n > 1)
```

If a dataset lacks primary keys, we can create a *surrogate key* by adding an ID column. A primary key and a foreign key form a *relation*. Relations are typically one-to-many.

### Mutating Joins

Let's create a smaller dataset.
```{r}
flights2 <- flights %>%
  select(year:day, hour, origin, dest, tailnum, carrier)
head(flights2)
```


If we want to add the airline names to `flights2`. We can do this with `left_join()`.
```{r}
flights2 %>%
  select(-origin, -dest)%>%
  left_join(airlines, by = "carrier")
```

We could accomplish the same thing using `mutate`.
```{r}
flights2 %>%
  select(-origin, -dest) %>%
  mutate(name = airlines$name[match(carrier, airlines$carrier)])
```

#### Inner Join

For an inner join, unmatched rows are not included in the result.

#### Outer Join

Keeps observations that appear in at least one of the tables. **Left join** keeps all observations in x, **right join** keeps all observations in y, and a **full join** keeps all observations in x and y. 

Left join is the most common and should be the default.

### Duplicate Keys

What happens if we our key isn't unique?
```{r}
x <- tribble(
  ~key, ~val_x,
     1,   "x1",
     2,   "x2",
     2,   "x3",
     1,   "x4"
)

y <- tribble(
  ~key, ~val_x,
     1,   "y1",
     2,   "y2",
)

left_join(x, y, by = "key")
```

With a duplicate key, all possible combinations are returned.

### Defining the Key Columns

- if we use `by = NULL`, all variables that appear in both tables are used to join the tables, also called a *natural* join

```{r}
flights2 %>%
  left_join(weather)
```


- join by a named character vector: `by = c("a"  = "b")`: variable `a` in table x matches to variable `b` in table y. The variables from x will be used in the output.

```{r}
flights2 %>%
  left_join(airports, c("dest" = "faa"))
```

```{r}
flights2 %>%
  left_join(airports, c("origin" = "faa"))
```

### Filtering Joins

Rather than affect variables, filtering joins affect the observations themselves.

- **semi_join(x, y)** *keeps* all observations in x that have a match in y
- **anti_join(x, y)** *drops* all observations in x that have a match in y

```{r}
top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest
```

We could filter this by hand:
```{r}
flights %>%
  filter(dest %in% top_dest$dest)
```

but once we add more variables it is much more difficult to filter by hand. We can use semi_join instead.
```{r}
flights %>%
  semi_join(top_dest)
```


Using anti-joins, we can easily find join mismatches.
```{r}
flights %>%
  anti_join(planes, by = "tailnum") %>%
  count(tailnum, sort = TRUE)
```




















