---
title: "Data Import with readr"
author: "Emma Grossman"
date: "4/11/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

`readr` turns files into data frames.

- `read_cvs()`: comma-delimited files
- `read_csv2()`: semicolon-separated files (common in countries where the comma is used as the decimal place)
- `read_tsv()`:tab-delimited files
- `read_delim()`: allows the user to specify any delimiter
- `read_fwf()`: fixed-width files
- `read_log()`: Apache style log files

We can supply an inline CSV file
```{r}
read_csv("a,b,c
         1,2,3
         4,5,6")
```

`read_csv()` is using the first line of data as column names. We can change this a couple ways:
```{r}
read_csv("The first line of metadata
         The second line of metadata
         x,y,z
         1,2,3", skip = 2)
```

Or by using comments:
```{r}
read_csv("# A comment I want to skip
         x,y,z
         1,2,3", comment = "#")
```

If the data do not have column names,
```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE)
```

We can also pass column names in as a vector:
```{r}
read_csv("1,2,3\n4,5,6", col_names =c("x","y","z"))
```

```{r error=TRUE}
read_csv("a,b\n1,2,3\n4,5,6")
read_csv("a,b\n\"1")
read_csv("a,b\n1,2\na,b")
read_csv("a;b\n1;3")
```

Parsing a Vector
```{r}
str(parse_logical(c("TRUE","FALSE","NA")))
str(parse_integer(c("1","2","3")))
str(parse_date(c("2010-01-01","1979-10-14")))
```

These parsers are the building blocks for `readr`

```{r}
parse_integer(c("1","241",".","456"), na = ".")
```

```{r error = TRUE}
x <- parse_integer(c("123","345","abc","123.45"))
```

```{r}
problems(x)
```

Parsing numbers is tricky because there is different notation in different countries.
```{r}
parse_double("1.23")
parse_double("1,23", locale = locale(decimal_mark = ","))
```

`parse_number` ignores non-numeric characters:
```{r}
parse_number("$100")
parse_number("20%")
parse_number("It cost $123.45")
```

```{r}
# Used in America
parse_number("$123,456,789")

# Used in many parts of Europe
parse_number(
  "123.456.789",
  locale = locale(grouping_mark = ".")
)

# Used in Switzerland
parse_number(
  "123'456'789",
  locale = locale(grouping_mark = "'")
)
```

Strings can also be difficult because there are different encodings. The most common now is UTF-8.

```{r}
charToRaw("Hadley")
charToRaw("Emma")
```

The remainder of subjects in this chapter are things that I know already, so I'm just going to copy the code without much explanation.

### Factors

```{r}
fruit <- c("apple","banana")
parse_factor(c("apple","banana", "bananana"), levels = fruit)
```



### Dates, Date-Times and Times

```{r}
parse_datetime("2010-10-01T2010")

# If time is omitted, it will be set to midnight
parse_datetime("20101010")

parse_date("2010-10-01")

library(hms)
parse_time("01:10 am")
parse_time("20:10:01")
```

```{r}
parse_date("01/02/15", "%m/%d/%y")
parse_date("01/02/15", "%d/%m/%y")
parse_date("01/02/15", "%y/%m/%d")

parse_date("1 janvier 2015", "%d %B %Y", locale = locale("fr"))
```


### Parsing a file

```{r}
guess_parser("2010-10-01")
guess_parser("15:01")
guess_parser(c("TRUE","FALSE"))
guess_parser(c("1","5","9"))
guess_parser(c("12,345,678"))
str(parse_guess("2010-10-10"))
```

### Problems

```{r}
challenge <- read_csv(readr_example("challenge.csv"))
```


```{r}
problems(challenge)
```


```{r}
challenge <- read_csv(
  readr_example("challenge.csv"),
  col_types = cols(
    x = col_double(),
    y = col_character()
  )
)

tail(challenge)
```

```{r}
challenge2 <- read_csv(
  readr_example("challenge.csv"),
  guess_max = 1001
)
```

```{r}
# read in data as character vectors

challenge2 <- read_csv(
  readr_example("challenge.csv"),
  col_types = cols(.default = col_character())
)
```

```{r}
df <- tribble(
  ~x, ~y,
  "1","1.21",
  "2","2.32",
  "3","4.56"
)
df
```

```{r}
type_convert(df)
```











