---
title: 'Case Study: k-NN and palmerpenguins'
author: "Emma Grossman"
date: "4/8/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(class)
library(gmodels)
```

## Import data
```{r}
penguins <- penguins %>%
  filter(!is.na(bill_length_mm))
```

## Normalize the continuous data

```{r}
normalize <- function(x){
  return ((x - min(x)) / (max(x) - min(x)))
}
```


```{r}
penguins_n <- as.data.frame(lapply(penguins[3:6], normalize))
```


## Create a testing and training data set

```{r}
set.seed(1223)
smp_size <- floor(0.75 * nrow(penguins))
train_id <- sample(1:nrow(penguins), smp_size)
penguins_train <- penguins_n[train_id,]
penguins_test  <- penguins_n[-train_id,]
```

Create label vectors of response categories:
```{r}
penguins_train_labels <- penguins[train_id, 1]
penguins_test_labels  <- penguins[-train_id, 1]
```

```{r}
sqrt(256)
```

Let's chose k = sqrt(256) = 16, since we have three levels of our response, a k of 16 will not result in a tie. 

```{r}
penguins_test_pred <- knn(train = penguins_train, test = penguins_test,
                           cl = penguins_train_labels[,1, drop = TRUE], 
                          k = 16) # wouldn't work without including [,1, drop = TRUE]
penguins_test_pred
```



Now to look at the classification table:
```{r}
CrossTable(x = penguins_test_labels[,1, drop = TRUE], y = penguins_test_pred,
           prop.chisq = FALSE) # wouldn't work without including [,1, drop = TRUE]
```

Only two penguins were misclassified!

## Try various k

Create a function to do this because why not? why shouldn't I?

```{r}
k_NN_dif_k <- function(k, seed){
  set.seed(seed)
  smp_size <- floor(0.75 * nrow(penguins))
  
  train_id <- sample(1:nrow(penguins), smp_size)
  penguins_train <- penguins_n[train_id,]
  penguins_test  <- penguins_n[-train_id,]
  
  penguins_train_labels <- penguins[train_id, 1]
  penguins_test_labels  <- penguins[-train_id, 1]
  
  penguins_test_pred <- knn(train = penguins_train, test = penguins_test,
                           cl = penguins_train_labels[,1, drop = TRUE], 
                          k = k)
  
  return(CrossTable(x = penguins_test_labels[,1, drop = TRUE], y = penguins_test_pred,
           prop.chisq = FALSE))
}
```

```{r}
k_NN_dif_k(k = 1, seed = 208)
```

So with k = 1, there were again, 2 penguins misclassified.

```{r}
k_NN_dif_k(k = 5, seed = 21035)
```

With k = 5, only 1 penguins was misclassified.

```{r}
k_NN_dif_k(k = 11, seed = 211)
```

Again, with k = 10 only 1 penguin was misclassified.
